Parsing Postfix log files

### 2007-11-22 doc/logparser-acronyms.tex
Finished the glossary entries for Postfix components.
Dealt with some more todo items and improved the other parsers section some more.
Further cleanup. Lots more XXX items. More work on logs2regexs documentation.

\newacronym{SPF}{Sender Policy Framework}{description={
    SPF is explained fully in~\cite{openspf, Wikipedia-spf}.
}}

@@

\newacronym{DFA}{Deterministic Finite Automata}{description={
    XXX EXPLAIN DFA.  SHOULD IT BE NFA?
}}

@@

>>>>

\newacronym{SPF}{Sender Policy Framework}{description={
    SPF is introduced in section~\refwithpage{spf introduction} and
    explained fully in~\cite{openspf, Wikipedia-spf}.
}}

@@

@@

\newacronym{LCD}{Lowest Common Denominator}{description={
    LCD is a mathematical term which is used figuratively to mean the least
    useful member of a set of alternatives.
}}

\newacronym{SLCT}{Simple Logfile Clustering Tool}{description={
    SLCT~\cite{slct-paper} is a tool implementing an algorithm designed by
    Risto Vaarandi~\cite{risto-vaarandi} for identifying and grouping
    similar log files, then producing a regex for each group of lines.
}}

### 2007-11-22 doc/logparser.tex
Ordered the actions alphabetically.
Further cleanup. Lots more XXX items. More work on logs2regexs documentation.

% When creating a PDF make the table of contents into links to the pages
% (without horrible red borders) and include bookmarks.  The title and
% author don't work - I think either gnuplot or graphviz clobbers it.
\usepackage{hyperref}
\hypersetup{
    pdftitle    = {Parsing Postfix log files},
    pdfauthor   = {John Tobin},
    final       = true,
    pdfborder   = {0 0 0},
%    breaklinks  = true,
}
\usepackage{breakurl}
\usepackage{lastpage}

@@

@@

Appendix~\ref{graphs} contains graphs illustrating the topics previously
discussed in section~\ref{rule efficiency}, rule efficiency, and explains
the irregularities observed in the graphs.

@@

\subsection{SMTP background}
\label{SMTP background}

The \SMTPlong, originally defined in \RFC{}~821~\cite{RFC821} and updated
in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
protocol, making it quite easy to test and debug problems with it.  Despite
the simplicity many virus and/or spam sending programs fail to implement it
properly, so requiring strict adherence to the protocol specification is
beneficial in protecting against spam and viruses.\footnote{This violates
the principle of \textit{Be liberal in what you accept, and conservative in
what you send\/} from \RFC{}~760~\cite{rfc760}, but unfortunately that
principle was written in a friendlier time.}  A typical \SMTP{}
conversation resembles the following (the lines starting with a three digit
number are sent by the server, all other lines are sent by the client):

@@

\subsection{Postfix background}

@@

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an \SMTP{} command, one list for each
stage of the \SMTP{} conversation: client connection, HELO command, MAIL
FROM command, RCPT TO command (possibly multiple times), DATA command, and
end of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter in Postfix's configuration controls
this behaviour.}  Postfix uses simple lookup tables as the deciding factor
for some restrictions, e.g.\
\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\footnote{\texttt{cidr}~\cite{cidr_table} is the type of the lookup table.}
will check whether the \IP{} address of the connected client matches the
left hand side of each line in the file and will return the right hand side
of the first matching line.  Other restrictions determine their result in
different ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com} checks
whether the \IP{} address of the client is present in the listed \RBL{}\@.

@@

    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.  \SPF{}
        records specify which mail servers are allowed to send mail
        claiming to be from a particular domain.  The intention is to
        reduce spam from faked sender addresses,
        backscatter~\cite{postfix-backscatter} and joe
        jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
        resistance to the proposal because it breaks or vastly complicates
        some features of \SMTP{}\@.

@@

\subsection{Other parsers}

@@

Parsing in \LMA{} is hand-coded and requires major changes to the code to
parse new log lines or extract additional data.  It does not appear to deal
with any of the complications discussed in
sections~\refwithpage{complications} and~\refwithpage{additional
complications}.  The program fails to parse Postfix 2.2.x or 2.3.x log
files; indeed it produces a number of warnings and no usable output.
Parsing in \parsername{} is divided in two (see
sections~\refwithpage{rules} and~\refwithpage{parsing-algorithm} for a more
detailed explanation): parsing a new log line is a trivial matter of adding
a new rule, but code changes may be required if new complications
(sections~\refwithpage{complications} and~\refwithpage{additional
complications}) must be dealt with or a new action is required
(section~\refwithpage{actions-in-detail}).

\LMA{} does provide some simple reports (IP-STORY, FROM-STORY, DAILY-EMAIL
and DAILY-REJECT); the author was unable to parse any logs with \LMA{}, so
didn't have an opportunity to review the reports.  \parsername{} does not
have any bundled reports, it delegates that task to a separate, dedicated
program.

In summary \LMA{} appears to be a proof of concept, written to demonstrate
the point of the paper (that having this information in an accessible
fashion is useful to systems administrators), rather than a program
designed to be extensible and useful in a production environment.

@@

        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
        \newline \hspace*{2em} server\_hostname} and \texttt{helo.\newline}
        \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
        \newline \hspace*{2em} pid\_regex, data} and \texttt{child.}

@@

            \item Create \texttt{/etc/postfix/log\_helo.pcre}
                containing:\newline
                \hspace*{2em}\texttt{/./~~~~WARN~Logging~HELO}

            \item Modify \texttt{smtpd\_data\_restrictions} in
                \texttt{/etc/postfix/main.cf} to contain\newline
                \hspace*{2em}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}

@@

The logs produced by Postfix differ from installation to installation, as
administrators have the freedom to choose the subset of available
restrictions which suits their needs, including using different \RBL{}
services, policy servers, or custom rejection messages.  To facilitate
this, the parser's design separates parsing rules from parsing actions:
adding new actions is difficult, but adding new rules to parse new
rejection messages is trivial (and occurs much more frequently).  The
implementation provides three programs to ease the process of adding new
rules: \texttt{add-new-rule}, \texttt{edit-rule} and \texttt{logs2regexs}
XXX NEED BETTER NAMES\@.

\texttt{add-new-rule} is the simplest of the three, it just starts the
user's preferred editor with a template rule which the user modifies as
required, and then inserts the new rule into the database.  The various
fields have already been described in sections~\refwithpage{rule
attributes},~\refwithpage{rule characteristics} and~\refwithpage{example
rule}, and will not be repeated here.

\texttt{edit-rule} extracts a rule from the database, saves it in a human
readable form, and runs the user's preferred editor for the user to change
the rule as required.  Once the user is satisfied with the changes, the
database will be updated appropriately.

\texttt{logs2regexs}

@@

@@

@@

    \item [SMTPD\_KILLED] Sometimes an event occurs which requires Postfix
        to forcefully kill a \daemon{smtpd}; that \daemon{smtpd} processes'
        active connection must be cleaned up, entered in the database, and
        deleted from the state tables.

    \item [BOUNCE] Postfix 2.3 (and hopefully subsequent versions) logs the
        creation of bounce messages.  This action creates a new mail if
        necessary, marking it as a bounce notification and removing the
        unknown origin flag if it exists.

@@

        AWStats will produce simple graphs for many different services, but
        supporting many different services without special purpose code
        restricts it to supporting the \LCD{}.  The data it will extract
        from an \MTA{} log file is limited to: \newline \hspace*{2em}
        time2, email, email\_r, host, host\_r, method, url, code and
        bytesd.  \newline  There does not seem to be an explanation of
        those fields in the documentation.  AWStats has no special purpose
        code to deal with the intricacies of Postfix logs, in fact it
        operates by coercing Postfix logs into Apache\footnote{The Apache
        web server is the most popular HTTP server in use for the past 10
        years; more information is available at
        \url{http://httpd.apache.org/}} format log files, for analysis by
        AWStats' HTTP log file parser.  The converting parser only deals
        with a small portion of the log lines generated by Postfix, and
        though it does correlate log lines by queueid, it does not deal
        with any of the other complications described in this document.

@@

        \hspace*{2em}Aggregations: count (not explained), bytes (sum of
        bytes transferred).\newline \hspace*{2em}Keyfields [sic]: date,
        hour, sender, recipient, defer\_count, delay, incoming\_host,
        outgoing\_host.

        YAALA provides an \SQL{} like mini-language used when generating
        the reports; sample reports can be seen at~\cite{yaala-samples}.
        Example query: \newline \hspace*{2em} \texttt{requests BY file
        WHERE host =\~{} Google}. \newline The mini-language is quite
        limited and cannot be used to extract data for external use, merely
        to create reports.

@@

@@

The mail loop was set up by a user modifying his mail forwarding to:
\newline \hspace*{2em}\texttt{$\backslash$username, username@domain}
\newline This instructs Postfix to deliver the mail to the local user, and
also forward it to the remote address; this is generally not a problem
except that the remote address in this case is the user's address, creating
an infinite loop.  To prevent this happening Postfix examines the
Delivered-To header in the mail, and if the mail has already been delivered
to the current address it is bounced back to the sender with the error
message \texttt{mail forwarding loop for username@domain}.  Ordinarily this
works well, but unfortunately in this case the user noticed they had not
received any mail in a while and opted to send a test mail to themselves,
causing a loop not caught by Postfix as described below.

@@

    \item GOTO 2.

>>>>

% When creating a \PDF{} make the table of contents into links to the pages
% (without horrible red borders) and include bookmarks.  The title and
% author don't work - I think either gnuplot or graphviz clobbers it.
\usepackage{hyperref}
\hypersetup{
    pdftitle    = {Parsing Postfix log files},
    pdfauthor   = {John Tobin},
    final       = true,
    pdfborder   = {0 0 0},
%    breaklinks  = true,
}
\usepackage{breakurl}
\usepackage{lastpage}

@@

\newcommand{\tab}[0]{%
    \hspace*{2em}%
}

@@

Appendix~\ref{graphs} contains graphs illustrating the topics previously
discussed in section~\ref{rule efficiency}, rule efficiency, and explains
the irregularities observed in these graphs.

@@

\subsection{Simple Mail Transfer Protocol}
\label{SMTP background}

The \SMTPlong{}, originally defined in \RFC{}~821~\cite{RFC821} and updated
in \RFC{}~2821~\cite{RFC2821}, is used for transferring mail between the
sending and receiving \MTA{}\@.  It is a simple, human readable, plain text
protocol, making it quite easy to test and debug problems with it.  Despite
the simplicity many virus and/or spam sending programs fail to implement it
properly, so requiring strict adherence to the protocol specification is
beneficial in protecting against spam and viruses.\footnote{This violates
the principle of \textit{Be liberal in what you accept, and conservative in
what you send\/} from \RFC{}~760~\cite{rfc760}, but unfortunately that
principle was written in a friendlier time. XXX CARL WANTS ME TO SUGGEST A
NEW PRINCIPAL.}  A typical \SMTP{} conversation resembles the following
(the lines starting with a three digit number are sent by the server, all
other lines are sent by the client):

@@

\subsection{Postfix}

@@

Postfix uses several restriction lists (each containing zero or more
restrictions) to decide the fate of an \SMTP{} command, one list for each
stage of the \SMTP{} conversation: client connection, HELO command, MAIL
FROM command, RCPT TO command (possibly multiple times), DATA command, and
end of data.  By default restriction lists will not be evaluated until the
first RCPT TO command is received, because some clients don't deal with
earlier rejections properly; a side benefit of this delay is that Postfix
has more information available when logging the rejection.\footnote{The
\texttt{smtpd\_delay\_reject} parameter in Postfix's configuration controls
this behaviour.}  Postfix uses simple lookup tables as the deciding factor
for some restrictions, e.g.\ XXX IMPROVE THIS EXPLANATION OR REMOVE IT \\
\tab{}\texttt{check\_client\_access~cidr:/etc/postfix/client\_access}
\footnote{\texttt{cidr}~\cite{cidr_table} is the type of the lookup table.}
\newline checks whether the \IP{} address of the connected client matches
the left hand side of each line in the file and will return the right hand
side of the first matching line.  Other restrictions determine their result
in different ways, e.g.\ \texttt{reject\_rbl\_client rbl.example.com}
checks whether the \IP{} address of the client is present in the listed
\RBL{}\@.

@@

    \item Checking \SPF{} records~\cite{openspf, Wikipedia-spf}.
        \SPF{}\label{spf introduction}
        records specify which mail servers are allowed to send mail
        claiming to be from a particular domain.  The intention is to
        reduce spam from faked sender addresses,
        backscatter~\cite{postfix-backscatter} and joe
        jobs~\cite{Wikipedia-joe-job}; however there has been a lot of
        resistance to the proposal because it breaks or vastly complicates
        some features of \SMTP{}\@.

@@

\subsection{Other Postfix log parsers}

@@

XXX IMPROVE THE NEXT PARAGRAPH BASED ON COMMENTS FROM CARL\@.

Parsing in \LMA{} requires major changes to the code to parse new log lines
or extract additional data.  It does not appear to deal with any of the
complications discussed in sections~\refwithpage{complications}
and~\refwithpage{additional complications}.  The program fails to parse
Postfix 2.2 or 2.3 log files; indeed all it produces is a number of
warning messages from the Perl interpreter, without any other output,
useful or otherwise.  Parsing in \parsername{} is divided in two (see
sections~\refwithpage{rules} and~\refwithpage{parsing-algorithm} for a more
detailed explanation): parsing a new log line is a trivial matter of adding
a new rule, but code changes may be required if new complications
(sections~\refwithpage{complications} and~\refwithpage{additional
complications}) must be dealt with or a new action is required
(section~\refwithpage{actions-in-detail}).

\LMA{} does provide some simple reports (IP-STORY, FROM-STORY, DAILY-EMAIL
and DAILY-REJECT); the author was unable to parse any logs with \LMA{}, so
didn't have an opportunity to review the reports.  \parsername{} does not
produce any reports, it delegates that task to a separate, dedicated
program.

In summary, \LMA{} appears to be a proof of concept, written to demonstrate
the point of their paper (that having this information in an accessible
fashion is useful to systems administrators), rather than a program
designed to be extensible and useful in a production environment.

@@

        \texttt{connection\_cols: client\_hostname, client\_ip, server\_ip,
        \newline \tab{} server\_hostname} and \texttt{helo.\newline}
        \texttt{result\_cols: sender, recipient, smtp\_code, message\_id,
        \newline \tab{} pid\_regex, data} and \texttt{child.}

@@

            \item Create \texttt{/etc/postfix/log\_helo.pcre}
                containing:\newline \tab{}\texttt{/./~~~~WARN~Logging~HELO}

            \item Modify \texttt{smtpd\_data\_restrictions} in
                \texttt{/etc/postfix/main.cf} to contain\newline
                \tab{}\texttt{check\_helo\_access~/etc/postfix/log\_helo.pcre}

@@

The logs produced by Postfix differ from installation to installation, as
administrators have the freedom to choose the subset of available
restrictions which suits their needs, including using different \RBL{}
services, policy servers, or custom rejection messages.  To facilitate
this, the parser's design separates parsing rules from parsing actions:
adding new actions is difficult, but adding new rules to parse new
rejection messages is trivial (and occurs much more frequently).  The
implementation provides a program to ease the process of creating new rules
from unparsed lines, based on the algorithm developed by Risto
Vaarandi~\cite{risto-vaarandi} for \SLCT{}~\cite{slct-paper}.  The
differences between the two algorithms will be outlines as part of the
general explanation below.

The core of the algorithm was developed by Risto Vaarandi, and is quite
simple: log lines are generally created by substituting variable words into
a fixed pattern, and analysis of the frequency with which each word occurs
determines whether the word is fixed or variable.  This classification can
be used to group similar log lines and generate a \regex{} to match each
group of lines.

There are 4 steps in the algorithm:

    \item [Pre-process the file]

        In the original algorithm the preprocessing stage was

    \item [Calculate the word frequency]

    \item [XXX]

    \item [Build regexs]

@@

    \item [BOUNCE] Postfix 2.3 (and hopefully subsequent versions) logs the
        creation of bounce messages.  This action creates a new mail if
        necessary; if the mail already exists the unknown origin flag will
        be removed.  The action also marks the mail as a bounce
        notification.

@@

    \item [SMTPD\_KILLED] Sometimes an \daemon{smtpd} is killed by a
        signal~\cite{Wikipedia-unix-signals}; the active connection for
        that \daemon{smtpd} must be cleaned up, entered in the database,
        and deleted from the state tables.

@@

@@

        AWStats will produce simple graphs for many different services, but
        supporting many different services without special purpose code
        restricts it to supporting the \LCD{}.  The data it will extract
        from an \MTA{} log file is limited to: \newline \tab{} time2,
        email, email\_r, host, host\_r, method, url, code and bytesd.
        \newline  There does not seem to be an explanation of those fields
        in the documentation.  AWStats has no special purpose code to deal
        with the intricacies of Postfix logs, in fact it operates by
        coercing Postfix logs into Apache\footnote{The Apache web server is
        the most popular HTTP server in use for the past 10 years; more
        information is available at \url{http://httpd.apache.org/}} format
        log files, for analysis by AWStats' HTTP log file parser.  The
        converting parser only deals with a small portion of the log lines
        generated by Postfix, and though it does correlate log lines by
        queueid, it does not deal with any of the other complications
        described in this document.

@@

        \tab{}Aggregations: count (not explained), bytes (sum of bytes
        transferred).\newline \tab{}Keyfields [sic]: date, hour, sender,
        recipient, defer\_count, delay, incoming\_host, outgoing\_host.

        YAALA provides an \SQL{} like mini-language used when generating
        the reports; sample reports can be seen at~\cite{yaala-samples}.
        Example query: \newline \tab{} \texttt{requests BY file WHERE host
        =\~{} Google}. \newline The mini-language is quite limited and
        cannot be used to extract data for external use, merely to create
        reports.

@@

XXX CAN I PUT \% ON THE Y AXIS\@?

@@

The mail loop was set up by a user modifying his mail forwarding to:
\newline \tab{}\texttt{$\backslash$username, username@domain} \newline This
instructs Postfix to deliver the mail to the local user, and also forward
it to the remote address; this is generally not a problem except that the
remote address in this case is the user's address, creating an infinite
loop.  To prevent this happening Postfix examines the Delivered-To header
in the mail, and if the mail has already been delivered to the current
address it is bounced back to the sender with the error message
\texttt{mail forwarding loop for username@domain}.  Ordinarily this works
well, but unfortunately in this case the user noticed they had not received
any mail in a while and opted to send a test mail to themselves, causing a
loop not caught by Postfix as described below.

@@

    \item GOTO 2.  XXX IMPROVE THIS, IT'S RUBBISH\@.
